Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2020-12-04 analysis/paper/paper.Rmd
re-organized to original outline

Habitat fragmentation and drought are major threats to biodiversity, and it has been hypothesized that they could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999]. Forest fragmentation is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008].

Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]. This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest. Drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001] <!--# not sure if this belongs here, for flow into next paragraph -->.

To date, however, this hypothesis remains untested for three primary reasons. First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants. <!--# example of juvenile but not reproductive study with drought -->. <!--# example of reproductive but not juvenile study with drought -->. Additionally, while most studies of tropical plant demography focus on trees, little is known about the understory plants that comprise up to \_\_\_\_% of biomass in tropical forests and are ecologically important <!--# citations -->.

1.  Organisms experience multiple sources of natural and anthropogenic stress simultaneously. However, the impacts of multiple sources of stress on population dynamics are not well understood [@tyeDemographicMenageTrois2016].

@@

    -   Especially difficult when lagged effects are expected. Part of the challenge in doing this modeling is the statistics are new <!--# maybe goes in discussion --> . "Move beyond the statistical shackles of interaction terms".LOL

2.  As a result of intense deforestation, tropical lowland forests are becoming an increasingly fragmented landscape. This is especially true in the Amazon basin where \_\_\_\_ km^2^ forest are within 100m of a forest edge <!--# citation -->.

    -   "Over 70000 km of new forest edges are being created annually in the Brazilian Amazon by human activities (Broadbent et al., 2008)"

    -   also cite: Kapos 1989, Lovejoy et al. 1986, Kapos et al. 1997?

    -   This can result in ...

3.  Simultaneously, these fragmented habitats are experiencing increased drought stress as a result of climate change. The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].

@@

    -   Combined environmental stresses, such as drought and fragmentation, have the potential to exacerbate or dampen effects on population dynamics [@tyeDemographicMenageTrois2016].

4.  Plant populations may be more susceptible to drought in fragmented forests. Fragmentation results in changes in forest microclimate that could exacerbate the effects of drought on plant populations. Near edges, tropical forests experience increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999].

    -   Trees near forest edges experience increased mortality after droughts compared to forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].

    -   Increased mortality after droughts disproportionately affects large trees.

@@

7.  We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape. Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest? 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest? 3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?

>>>>

1.  Organisms experience multiple sources of natural and anthropogenic stress simultaneously. However, the impacts of multiple sources of stress on population dynamics are not well understood. Combined environmental stresses have the potential to interact to exacerbate or dampen additive effects on population dynamics. <!--# examples of interactions on demography -->For example, drought and habitat fragmentation ...[@tyeDemographicMenageTrois2016].

@@

2.  Habitat fragmentation is one such source of environmental stress. Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest farther than 1 km from an edge has increased to 75% compared to 90% historically [@haddadHabitatFragmentationIts2015]. Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->

3.  Simultaneously, fragmented tropical forests are experiencing increased drought stress as a result of climate change. The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].

@@

4.  It has been hypothesized that drought and fragmentation could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999]. Fragmentation results in changes in forest microclimate near edges including increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]. This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest. For example, drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].

@@

<!--# transition better -->To date, however, this hypothesis remains untested for three primary reasons. First, most studies elucidating how tropical plants respond to fragmentation have studied either juvenile (i.e., seed, seedling) or larger, reproductive plants. <!--# example of juvenile but not reproductive study with drought -->. <!--# example of reproductive but not juvenile study with drought -->. Additionally, while most studies of tropical plant demography focus on trees, little is known about the understory plants that comprise up to \_\_\_\_% of biomass in tropical forests and are ecologically important <!--# citations -->.

<!--# more about the heliconia dataset here -->

We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape. Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest? 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest? 3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?
